---
title: A little math can go a long way
subtitle: >
  This document explores methods to calculate the sum of integers ≤ N divisible by 3 or 5. It includes R implementations, performance comparisons, and mathematical optimizations.
date: 2024-01-15
categories: [programming, optimizing-code]
execute: 
  cache: true
---

I'm exploring the following problem as a potential exercise for students in my Scientific Computing course:

::: {.callout-note appearance="simple"}
## Problem

Calculate the sum of all positive integers smaller or equal to N which are divisible by 3 or 5. For example, if N = 20, then the answer is 3+5+6+9+10+12+15+18+20 = 98
:::

```{r setup}
#| include: false
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(autodep = TRUE)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
```

```{r testing-fun}
#| code-fold: true
#| code-summary: Unit tests
test <- function(.f) {
  structure(
    list(
      tinytest::expect_equal(.f(3), 3, info = "Input: N = 3"),
      tinytest::expect_equal(.f(-100), 0, info = "Input: N = -100"),
      tinytest::expect_equal(.f(20), 98, info = "Input: N = 20"),
      tinytest::expect_equal(.f(6), 14, info = "Input: N = 6"),
      tinytest::expect_equal(.f(0), 0, info = "Input: N = 0")
    ),
    class = "tinytests"
  )
}
```

## Code smell

Here is a horrible way to implement this in R:

```{r}
sum_multiples_loop <- function(N) {
  out <- 0
  for (i in 1:N) {
    if (i %% 3 == 0 || i %% 5 == 0) {
      out <- out + i
    }
  }
  return(out)
}

sum_multiples_loop(20)
```

It works, but there are a ton of problems that make this code really slow and we can make it several thousand times faster by the end of this post. But it is also a good idea to test more than one case, and since I don't want to copy paste calls to each function for many different scenarios, I have written a small suite of unit tests:

```{r}
test(sum_multiples_loop)
```

Not surprisingly, our function fails when the input is less than 0, so this is something that we need to address later.

## Standard vectorization

We can of course vectorize a lot and simplify the whole thing dramatically by using standard R vectorization. Instead of looping over positive integers from 1 to N, we create a vector x which stores those integers. Then we take advantage some vectorizations

```{r}
sum_multiples_vectorized <- function(N) {
  if (N <= 0) return(0)                         # <1>
  x <- seq_len(N)                               # <2>
  is_divisible <- x %% 3 == 0 | x %% 5 == 0     # <3>
  sum(x[is_divisible])                          # <4>
}

test(sum_multiples_vectorized)
```

1.  Take care of edge cases with an early return; alternative is to give an error
2.  seq_len(N) is a safer alternative to 1:N (doesn't matter here because we return early if N is not-positive, but good practice in general)
3.  Boolean operations on vectors produce vectors of TRUE and FALSE values, so no need for loops and if statements
4.  Subseting a vector with a boolean vector will return a new vector with only those values for which `is_divisible` is TRUE; `sum` is vectorized and will return the sum of all elements in x

We can now compare the performance of the two functions. There are many ways to do this. Here is one popular option:

```{r}
bench::mark(
  sum_multiples_loop(1e6),
  sum_multiples_vectorized(1e6),
  relative = TRUE
)
```

but we can do the math and get much better results:

```{r}
sum_multiples_math <- function(N) {
  if (N <= 0) {
    return(0)
  }
  triangular_number <- function(n) {
    n * (n + 1) / 2
  }

  a <- 3
  b <- 5
  c <- 15

  a * triangular_number(floor(N / a)) +
    b * triangular_number(floor(N / b)) -
    c * triangular_number(floor(N / c))
}

test(sum_multiples_math)
```

compare to the other two functions:

```{r}
bench::mark(
  sum_multiples_loop(1e6),
  sum_multiples_vectorized(1e6),
  sum_multiples_math(1e6),
  relative = TRUE
)
```

of course, in all cases we can parameterize the function rather than hard-coding it. The challenge would be to find the least common multiple. You could use the function `pracma::Lcm`

```{r}
sum_series_divisors <- function(N, a, b) {
  if (N <= 0) {
    return(0)
  }
  triangular_number <- function(n) {
    n * (n + 1) / 2
  }

  lcm <- pracma::Lcm(a, b)

  a * triangular_number(floor(N / a)) +
    b * triangular_number(floor(N / b)) -
    lcm * triangular_number(floor(N / lcm))
}

test(\(x) sum_series_divisors(x, a = 3, b = 5))
```

which incurs a cost of course, but that cost is minuscule relative to the other approaches above:

```{r}
sum_series_divisors_3_5 <- \(x) sum_series_divisors(x, a = 3, b = 5)
bench::mark(
  sum_multiples_math(1e6),
  sum_series_divisors_3_5(1e6),
  relative = TRUE
)
```

but let's not use any libraries and write our own lcm function
